---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# misc

<!-- badges: start -->
<!-- badges: end -->

## Installation

You can install the released version of misc from [github](https://github.com/kguidonimartins/misc) with:

```{r, message=FALSE}
if (!require("remotes")) install.packages("remotes")
if (!require("bdc")) remotes::install_github("brunobrr/bdc")
```

## Introducing BDC

**BDC - a comprehensive and straightforward workflow for integrating, standardizing, and cleaning biodiversity data**

Handle biodiversity data from several varying sources is not an easy task. This workflow was created to facilitate the process of integrating, standardizing, and cleaning biodiversity data aiming to improve its and make it fit for use.

The workflow is composed of five steps:

1.  Merge datasets: Standardization and integration of datasets from several heterogeneous sources
2.  Prefilter: flagging, identifying, and cleaning data missing or with invalid information
3.  Taxonomy: cleaning and standardizing scientific names (ten taxonomic authorities available and fuzzy match option allowed)
4.  Space: flagging, identifying, and cleaning potentially wrong geographic coordinates (fifteen tests available)
5.  Time: standardizing and flagging information about an event

Aim to make easier the interpretation and visualization of results, in each step of the workflow a report and figures are created. Further, standardized databases resulting from each step of the workflow as well as databases containing information that needs (could) to be checked by users are automatically saved.
