---
title: "Installing gnparser"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing gnparser}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo = TRUE,
  warning = FALSE,
  eval = F
)
```

This tutorial will guide you through the installation of `gnparser`, a required binary file for using our R package. The installation process varies depending on your operating system (Windows, macOS, or Linux).

We provide two installation options: using reproducible code or following a manual step-by-step installation. The first option is a quick and easy way to install the software. In case of errors, you can try the manual step-by-step approach.

### **Option 1: Run the following code**

Installing and loading bdc package

```{r}
install.packages("bdc")
library(bdc)
```

<br/>

⚠ **Important**: Get the latest version of gnparser. The execution of this step is required independently of our operation system.

```{r}
  get_latest_gnparser_version <- function() {
    url <- "https://github.com/gnames/gnparser/releases"
    
    # Reads the content of the releases page
    page <- xml2::read_html(url)
    
    # Extracts the first tag containing the version, using a CSS selector
    version <- page %>%
      rvest::html_element(".Link--primary") %>%
      rvest::html_text2()
    
    version <- gsub("v", "", version)
    return(version)
  }

  latest_version <- get_latest_gnparser_version()
  message("Latest version of gnparser: ", latest_version)
```

#### **Windows**

**Download the lastest version of gnparser**

```{r}
link <- "https://github.com/gnames/gnparser/releases/download/"
url_latest_version = paste0(link, "v", latest_version, "/gnparser-v", latest_version, "-win-arm.zip")
file_name = here::here(paste0(latest_version, ".zip"))
                      
download.file(url_latest_version, file_name)
```

**Move file to the recommended directory**

```{r}
# Extracts the file in a temporary directory
unzip(file_name, exdir = here::here())

# Sets the destination filename
AppData_path <- Sys.getenv("AppData")

# Moves the executable to the target directory
file.copy("gnparser.exe", AppData_path, recursive = TRUE)
```

<br/>

#### **Linux**

**Download the lastest version of gnparser**

```{r}
link <- "https://github.com/gnames/gnparser/releases/download/"
url_latest_version = paste0(link, "v", latest_version, "/gnparser-v", latest_version, "-linux-arm.tar.gz")
file_name = here::here(paste0(latest_version, ".tar.gz"))
                      
download.file(url_latest_version, file_name)
```

**Move file to the recommended directory**

```{r}
# Extracts the file in a temporary directory
untar(file_name, exdir = here::here())

# Moves the executable to the target directory
file.copy(here::here("gnparser"), "~/bin")
```

<br/>

#### **macOS**

⚠️**IMPORTANT:** Before installing gnparser, it is highly recommend to install Homebrew, which is a package manager for macOS and Linux that simplifies the installation of software and libraries directly from the terminal. You can use the steps below to install Homebrew. Moreover, a detailed explanation on how to install homebrew can be found in [this video](https://www.youtube.com/watch?v=IWJKRmFLn-g&ab_channel=EasyOSX).

1.  **Open Terminal**

    -   Press `Command + Space` to open Spotlight Search
    -   Type "Terminal" and press Enter
    -   A new window will appear - this is your Terminal

2.  **Using Homebrew (Recommended)**

    -   In the Terminal window, paste and run these commands:

``` bash
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
```

3.  **Install gnparser** **using brew**

``` bash
brew tap gnames/gn 
brew install gnparser
```

After that, close the terminal and **refresh the R section**.

<br/>

### **Option 2: Manual Installation**

#### **Windows**

1.  **Download the installer**

    -   Go to [gnparser releases](https://github.com/gnames/gnparser/releases)
    -   Download the Windows version (file ending in `-win-64.zip`)
    -   Save it to a known location on your computer

2.  **Open Command Prompt**

    -   Press `Windows + R` on your keyboard
    -   Type `cmd` and press Enter
    -   A black window will appear - this is your Command Prompt

3.  **Create installation directory**

    -   In the Command Prompt window, type:

    ``` cmd
    mkdir C:\bin
    ```

4.  **Extract and move the file**

    -   Extract the downloaded zip file

    -   Copy the `gnparser.exe` file to C:\bin using either:

        -   Windows Explorer (drag and drop)
        -   Or Command Prompt:

        ``` cmd
        copy path\to\downloaded\gnparser.exe C:\bin
        ```

5.  **Add to System Path**

    -   Press `Windows + X` and select "System"
    -   Click "Advanced system settings"
    -   Click "Environment Variables"
    -   Under "System variables", find and select "Path"
    -   Click "Edit"
    -   Click "New"
    -   Add `C:\bin`
    -   Click "OK" on all windows

6.  **Verify installation**

    -   Open a new Command Prompt window
    -   Type:

    ``` cmd
    gnparser --version
    ```

    -   If you see a version number, the installation was successful

<br/>

#### **Linux**

1.  **Open Terminal**

    -   Press `Ctrl + Alt + T` (on most Linux distributions)
    -   Or find Terminal in your applications menu
    -   A new window will appear - this is your Terminal

2.  **Download the binary**

    -   In the Terminal window, run these commands:

    ``` bash
    # Create a temporary directory
    mkdir -p ~/tmp
    cd ~/tmp

    # Download the latest Linux version
    wget https://github.com/gnames/gnparser/releases/download/v1.10.1/gnparser-v1.10.1-linux.tar.gz

    # Extract the file
    tar xvf gnparser-*.tar.gz

    # Move to system directory
    sudo mv gnparser /usr/local/bin
    ```

3.  **Verify installation**

    `bash gnparser --version`

<br/>

#### **macOS**

1.  **Open Terminal**

    -   Press `Command + Space` to open Spotlight Search
    -   Type "Terminal" and press Enter
    -   A new window will appear - this is your Terminal

2.  **Using Homebrew (Recommended)**

    -   In the Terminal window, paste and run these commands:

    ``` bash
    # Install Homebrew if you haven't already
    /bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"

    # Install gnparser
    brew tap gnames/gn
    brew install gnparser
    ```

3.  **Manual Installation**

    -   Download the macOS version (file ending in `-mac.tar.gz`) from [gnparser releases](https://github.com/gnames/gnparser/releases)
    -   Open Terminal (Applications \> Utilities \> Terminal)
    -   Navigate to the download directory:

    ``` bash
    cd ~/Downloads
    ```

    -   Extract the file:

    ``` bash
    tar xvf gnparser-v*.tar.gz
    ```

    -   Move to system directory:

    ``` bash
    sudo mv gnparser /usr/local/bin
    ```

    -   You may need to allow the application in System Preferences:
        -   Go to System Preferences \> Security & Privacy
        -   Click the lock to make changes
        -   Allow gnparser to run \`\`\`

<br/>

### **Troubleshooting**

If you encounter any issues:

1.  **Command not found**

    -   Make sure you've added the correct directory to your system PATH
    -   Try closing and reopening your terminal/command prompt

2.  **Permission denied**

    -   On macOS/Linux, you may need to make the file executable:

    ``` bash
    chmod +x /usr/local/bin/gnparser
    ```

3.  **Security warnings**

    -   On macOS: Go to System Preferences \> Security & Privacy and allow the application
    -   On Windows: You might need to run as administrator

### **Testing the Installation**

After installation, test if gnparser is working correctly:

```{r}
# In R, try parsing a scientific name
test <-
  bdc::bdc_clean_names(c("OCOTEA odorifera", "ocotea odorifera", "ocotea cf. odorifera"))
```
